A Decision Procedure for (Co)datatypes in SMT Solvers

### 2014-12-02 tex/codatadp.tex
more work on intro etc.

\begin{abstract}
Codatatypes naturally capture potentially infinite data structures and
processes. We introduce a decision procedure that combines reasoning about
datatypes and codatatypes. The dual of the acyclicity rule for datatypes is a
bisimilarity rule that identifies observationally equal, possibly cyclic
values. The procedure is compatible with the Nelson--Oppen method
and is useful for both proving and model finding. It has been implemented in the
latest version of CVC4, a modern SMT solver. An evaluation based on problems
generated from Isabelle/HOL theories demonstrates the potential of the procedure
in a proving context.
\end{abstract}

@@

  * briefly: datatypes and why they are useful
    (terminology: freely-generated, inductive, algebraic, ..., sometimes
    with different meanings; we'll clarify below)
    * Hoare's recursive data structures
    * comparatively easy to reason about and automate
    * common in practice, esp. in computer science applications

@@

\paragraph{Notations.}
Blah

@@

@@

\newcommand\keyw[1]{\textbf{#1}}
\newcommand\const[1]{\textsf{#1}}
\newcommand\ty[1]{\mathit{#1}}

For the purposes of this paper, a (co)datatype specification consists of
$l$~mutually recursive types which are either all datatypes or all codatatypes.
Each datatype $\delta$ is equipped by $m \ge 1$ constructors, and each
constructor for $\delta$ takes $n \ge 0$ arguments, each associated with a
selector, and returns a $\delta$ value. The argument types must be either
already existing types or be one of the newly introduced (co)datatypes. The
introduced names for the (co)datatypes, the constructors, and the selectors must
be distinct with already introduced ones and among themselves. Schematically:
%
\[
\begin{aligned}[t]
(\keyw{co})\keyw{datatype}\;\,
  \delta_1 & {}=\, \const{C}_{11}(\bigl[\const{s}_{111}{:}\bigr]\; \sigma_{111}, \,\ldots,\, \bigl[\const{s}_{11n_{11}}{:}\bigr]\; \sigma_{11n_{11}}) & \\
           & {}\;\phantom{=\,}\;{} \smash{\,\,\vdots} \\ % {}\;\phantom{=}\llap{\ensuremath{\mid}}\; \cdots \\
           & {}\;\phantom{=\,}\llap{\ensuremath{\mid}}\; \const{C}_{1m}(\bigl[\const{s}_{1m1}{:}\bigr]\; \sigma_{1m1}, \,\ldots,\, \big[\const{s}_{1mn_{1m}}{:}\big]\; \sigma_{1mn_{1m}}) \\[-1\jot]
   \smash{\vdots\,\,\,} \\[-1\jot]
  \keyw{and}\; \,\delta_l & =\, \ldots
\end{aligned}
\]
%
with
$\const{C}_{ij} : \sigma_{ij1}\times\cdots\times\sigma_{ijk_{ij}} \to \delta_i$
and $\const{s}_{ijk} : \delta_i \to \sigma_{ijk}$. The selector names are often
omitted in our examples, in which case default names are generated.

Polymorphic types, nested (co)recursion, and datatype--codatatype mixtures fall
outside this fragment. For example, the specification
\[\begin{aligned}[t]
      \keyw{datatype}\;\, \ty{tree} & \,=\, \const{Node}(\ty{int},\, \ty{forest}) \\[-\jot]
      \keyw{and}\;\, \ty{forest} & \,=\, \const{FNil} \mid \const{FCons}(\ty{tree},\, \ty{forest})
\end{aligned}
\]
is supported, whereas
\[\begin{aligned}[t]
      \keyw{codatatype}\; \,\ty{llist}(\alpha) & \,=\, \const{LNil} \mid \const{LCons}(\alpha,\, \ty{llist}(\alpha)) \\
      \keyw{datatype}\;\, \ty{tree} & \,=\, \const{Node}(\ty{int},\, \ty{llist}(\ty{tree}))
\end{aligned}
\]
is beyond the fragment considered here.
% (although it makes sense and is supported by Isabelle \cite{xxx}).

@@

  * Constructed examples
  * Isabelle benchmarks

>>>>

\begin{abstract}
Codatatypes naturally capture potentially infinite data structures and
processes. We introduce a decision procedure that combines reasoning about
datatypes and codatatypes. The dual of the acyclicity rule for datatypes is a
bisimilarity rule that identifies observationally equal, possibly cyclic values.
The procedure is complete for the universal fragment and is compatible with the
Nelson--Oppen method. It has been implemented in the latest version of CVC4, a
modern SMT solver. An evaluation based on problems generated from Isabelle
theories demonstrates the potential of the procedure for both proving and model
finding.
\end{abstract}

@@

  * briefly: datatypes and why they are useful
    * Hoare's recursive data structures
    * comparatively easy to reason about and automate
    * common in practice, esp. in computer science applications

    (terminology: freely-generated, inductive, algebraic, ..., sometimes
    with different meanings; we'll clarify below what we use)

@@

@@

\paragraph{Conventions.}

* types are simply atomic sorts, with no structure, and interpreted by nonempty
  domains

* although nothing prevents composing the decision procedure with theories
  providing polymorphic types (parametric sorts), such as the theory of arrays

* $\sigma, \tau$ range over types---whether (co)datatypes or uninterpreted types
* $\delta$ range over (co)datatypes

* use sans-serif for function symbols (e.g., $\const{f}$) to distinguish them
from variables (e.g., $x$).

* function symbols are accompanied by a type signature, e.g. $\const{f} : \sigma_1 \times \cdots \times \sigma_n \to
\sigma$ indicates

* don't distinguish between constants and function symbols
  a constant is merely the case $n = 0$ of a function $f :

@@

\newcommand\keyw[1]{\textbf{#1}}
\newcommand\const[1]{\textsf{#1}}
\newcommand\ty[1]{\textit{#1}}

A signature consists of:
  * set of plain types, which can be interpreted or uninterpreted
    * must contain the distinguished \ty{bool} type of Booleans
  * set of function symbols with argument and return types
  * list of codatatype specifications, described below

A (co)datatype specification consists of $l$~mutually recursive types which are
either all datatypes or all codatatypes. Each datatype $\delta$ is equipped by
$m \ge 1$ constructors, and each constructor for $\delta$ takes zero or more
arguments and returns a $\delta$ value. The argument types must be either
plain, be or be one of the newly introduced (co)datatypes. To every
argument corresponds a selector. The introduced names for the (co)datatypes, the
constructors, and the selectors must be distinct and different from already
introduced ones. Schematically:
%
\[
\begin{aligned}[t]
(\keyw{co})\keyw{datatype}\;\,
  \delta_1 & {}=\, \const{C}_{11}(\bigl[\const{s}_{111}{:}\bigr]\; \sigma_{111}, \,\ldots,\, \bigl[\const{s}_{11n_{11}}{:}\bigr]\; \sigma_{11n_{11}}) & \\
           & {}\;\phantom{=\,}\;{} \smash{\,\,\,\vdots} \\[-\jot] % {}\;\phantom{=}\llap{\ensuremath{\mid}}\; \cdots \\
           & {}\;\phantom{=\,}\llap{\ensuremath{\,\mid\,}}\; \const{C}_{1m}(\bigl[\const{s}_{1m1}{:}\bigr]\; \sigma_{1m1}, \,\ldots,\, \big[\const{s}_{1mn_{1m}}{:}\big]\; \sigma_{1mn_{1m}}) \\[-1\jot]
   \smash{\vdots\,\,\,} \\[-1\jot]
  \keyw{and}\; \,\delta_l & =\, \ldots
\end{aligned}
\]
%
with
$\const{C}_{ij} : \sigma_{ij1}\times\cdots\times\sigma_{ijk_{ij}} \to \delta_i$
and $\const{s}_{ijk} : \delta_i \to \sigma_{ijk}$. Defaults are assumed for
the selector names if they are omitted.

Polymorphic types, nested (co)recursion, and datatype--codatatype mixtures fall
outside this fragment. For example, the specification
\[\begin{aligned}[t]
      \keyw{datatype}\;\, \ty{tree} & \,=\, \const{Node}(\ty{int},\, \ty{forest}) \\[-\jot]
      \keyw{and}\;\, \ty{forest} & \,=\, \const{FNil} \,\mid\, \const{FCons}(\ty{tree},\, \ty{forest})
\end{aligned}
\]
is supported, whereas
\[\begin{aligned}[t]
      \keyw{codatatype}\; \,\ty{llist}(\alpha) & \,=\, \const{LNil} \,\mid\, \const{LCons}(\alpha,\, \ty{llist}(\alpha)) \\[-.5\jot]
      \keyw{datatype}\;\, \ty{tree} & \,=\, \const{Node}(\ty{int},\, \ty{llist}(\ty{tree}))
\end{aligned}
\]
is beyond the fragment considered here. In principle, rank-1 (top-level)
polymorphism would be easy to support; nesting datatypes inside datatypes,
and likewise for codatatypes, can be reduced to the mutual case. So the only
genuinely interesting cases missing are mixed nested (co)recursion
(as in the \ty{tree} example above) as well as (co)recursion through a
non-(co)datatype (both of which do make sense \cite{blanchette-et-al-2014-codata}).

@@

  * Constructed examples
    * proving
    * model finding?

  * Isabelle benchmarks
